Defining Transcriptome using RNA Sequencing\\ of {\em Streptococcus mutans}

### 2012-01-21 doc/manuscript.tex
Just and update.

@@

>>>>

\fixme{I might have to try different filters of short reads to see how those
affect the transcript prediction.}

@@

The input RNA-seq data is a list of site-wise non-negative integers that
represents a number of reads mapped to a particular genomic position
\citet{Martin2010}; a RNA-seq ``coverage'' data. We designated a gene as an
expressed if the gene belongs to a transcript with at least C2 level.

The number of transcripts predicted was

A gene

We use the term `coverage' to designate 
We verified the genes using RNA-seq data.

### 2012-01-21 package/manuscript.Rnw
Just and update.

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color} 
\usepackage{setspace} 
\usepackage{rotating}
\usepackage{natbib}
\usepackage{xr}
\usepackage{../../latex/sty/genres}

@@

\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=6,height=6}
\newcommand{\Rpackage}[1]{R package \textit{#1}}
\newcommand{\program}[1]{{\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}

@@

\author{Sang Chul Choi$^{1}$, Charles D.\ Danko$^{1}$, 
Melissa J.\ Hubisz$^{1}$, \\
Robert A.\ Burne$^{3}$
Michael J.\ Stanhope$^{2}$, and Adam Siepel$^{1}$}

@@

\section{Abstract}
A bacterial transcriptome from the primary etiological agent of human dental
caries, \textit{Stretptococcus mutans}, is described using deep RNA sequencing
data.  We define transcripts boundaries of open reading frames based on sequence
alignments of short reads on the reference genome of UA159 from the deep
sequencing.  Based on homologous sequences and RNA secondary structures we also
report putative small RNAs, several of which are confirmed experimentally using
both RNA sequencing data, northern blot, and qRT-PCR.  CcpA is known to play a
direct role in carbon catabolite repression (CCR).  We use the developed
transcript map, open read frame, and small RNAs to determine differentially
expressed regions between the wild-type (UA159) and CcpA-deletion (TW1) strains
in order to refine the previously established importance of CcpA regulation as
central metabolism and virulence gene expression.  We find that more genes are
differentially expressed in the comparison between wild-type and CcpA-deletion
strains that are grown in galactose than in glucose.  We also find that more
genes differentially expressed in wild-type than in CcpA-deletion strain when
comparing glucose- and galactose-grown cells.  These two findings are different
from those from microarry experiment of \cite{Abranches2008}.  We also predicted
target genes of the small RNAs that were enriched in the following functional
categories: xxx.  We confirme that CcpA modulates the pathogenic potential of
\textit{S.\ mutans} through global control of gene expression.

\section{Abbreviations}

\begin{enumerate}
\item CCR Carbon catabolite repression
\item CcpA ?
\item IPS intracellular polysaccharide
\item CRE catabolite responsive elements
\end{enumerate}

@@

Although our RNA-seq data did not sampled specifically non-coding small RNAs, we
retained small RNAs in the sample by attempting to deplete only ribosomal RNAs
from the sample. Because non-coding small RNAs and coding messenger RNAs were
mixed in the sample, we sought a way of predicting small RNAs using
\program{RNAz} developed by \citet{Gruber2010}.  \program{RNAz} used homologous
sequences and RNA secondary structures to predict putative non-coding RNAs. We
confirmed the several predicted small non-coding RNAs experimentally
\citep{Altuvia2007} including the RNA-seq data set. Because small RNAs could
be associated with CCR regulation of genes (CITE), we determined differentially expressed small
RNAs between RNA samples. We found a few small RNAs to be differentially
expressed.

Bacteria can selectively use substrates from different carbon sources. In the
presence of a preferred carbon source, catabolic system of secondary substrates
are often regulated or shutdown: called carbon catabolite repression (CCR)
\cite{Gorke2008}. CCR also regulates the expression of virulence factors in many
pathogenic bacteria. A classical example is the glucose-lactose diauxie in
\textit{E.\ coli}. A significant number of genes can be subject to CCR.  CCR in
bacteria involves regulatory networks that activate or silence genes in response
to carbon source and availability.  \textit{S.\ mutans} possesses redundant
systems for CCR.  We could measure directly the expression levels of
\textit{fruA} in \textit{S.\ mutans}.  We assess the contribution of CcpA to CCR
of predicted operons.  Clearly, deletion of \textit{ccpA} resulted in
substantial relief of the CCR of \textit{fruA} expression exerted by both
glucose and fructose.

@@

We need to quantify some uncertainty of transcript predictions.

@@

\subsection{Bacterial strains and growth conditions}
\fixme{Burne group should be able to describe this part. The following is just
from somewhere for the purpose of writing template.}
\textit{S.\ mutans} UA159 is a human cariogenic isolate. The whole genome is
availabe at NCBI \cite{Ajdic2002}. Strain TW1 was derived from UA159 by
replacing a part of \textit{ccpA} gene \cite{Wen2002}.  \textit{S.\ mutans}
strains were inoculated into TV medium supplemented with 0.5\% galactose or
glucose and grown to early exponential phase (OD600, 0.2).  Then fructose was
added to the medium at concentrations (0,0.05,3,5,10 mM) for 3.5h to induce
\textit{fruA} expression.

\subsection{RNA isolation}
\fixme{Burne group should be able to describe this part. The following is just
from somewhere for the purpose of writing template.}
RNA was isolated from S. mutans (1) by using 50 ml cultures that were grown
under the desired conditions and harvested by centrifugation at $4^{\circ}$C. Pelleted
cells were resuspended in 400 ul of diethyl pyrocarbonate-treated water, 800 ul
of RNAprotect reagent (Qiagen, Inc., Chatsworth, CA) was added, and the samples
were incubated at room temperature for 5 min with vortexing for 10 s at 1-min
intervals. Cells were then pelleted, resuspended in 250 ul Tris-EDTA (50:10)
buffer, and transferred to 1.5 ml screw-cap tubes containing sterile glass beads
(average diameter, 0.1 mm; Biospec, Bartlesville, OK), 10 ul of 1\% sodium
dodecyl sulfate, and 300 ul of acid phenol-chloroform (5:1). After
centrifugation for 15 min at maximum speed at $4^{\circ}$C, RNA was further purified
using an RNeasy mini kit (Qiagen, Inc., Chatsworth, CA), including on-column
DNase digestion with RNase-free DNase (Qiagen), as recommended by the supplier.

\subsection{RNA sequencing}
\fixme{Stanhope group should be able to describe this part. The following is just
from somewhere for the purpose of writing template.}
Sequencing libraries for the Illumina GA platform were constructed by shearing
the enriched cDNA by nebulisation (35psi, 6 min) followed by end-repair with
Klenow polymerase, T4 DNA polymerase and T4 polynucleotide kinase (to blunt-end
the DNA fragments). A single 3' adenosine moiety was added to the cDNA using
Klenow exo- and dATP. The Illumina adapters (containing primer sites for
sequencing and flowcell surface annealing) were ligated onto the repaired ends
on the cDNA and gel-electrophoresis was used to separate library DNA fragments
from unligated adapters by selecting cDNA fragments between 200-250 bps in size.
Fragmentation followed by gel-electrophoresis were used to separate library DNA
fragments and size fragments were recovered following gel extraction at room
temperature to ensure representation of AT rich sequences. Ligated cDNA
fragments were recovered following gel extraction at room temperature to ensure
representation of AT rich sequences. Libraries were amplified by 18 cycles of
PCR with Phusion polymerase. Sequencing libraries were denatured with sodium
hydroxide and diluted to 3.5 pM in hybridisation buffer for loading onto a
single lane of an Illumina GA flowcell. Cluster formation, primer hybridisation
and single-end, 36 cycle sequencing were performed using proprietary reagents
according to manufacturers' recommended protocol (https://icom.illumina.com/).
The efficacy of each stage of library construction was ascertained in a quality
control step that involved measuring the adapter-cDNA on a Agilent DNA 1000
chip. A final dilution of 2 nM of the library was loaded onto the sequencing
machine.

SOLiD sequencing was performed at Agencourt Bioscience (Beverly, MA) and
SeqWright (Houston, TX). Library preparations, fragment library protocols, and
all SOLiD-run parameters followed standard Applied Biosystems protocols.
Illumina/Solexa sequencing followed standard Illumina/Solexa methods and
protocols. Both systems were used because in this study our aim was simply to
collect as much sequence coverage as possible, and in pursuing this goal, we
utilized all available sequencing systems. We note, however, that the relative
performances (overall coverage, data output, etc.) noted in Table 1 for the
Illumina and Applied Biosystems sequencing platforms cannot be compared
directly. This study was not intended or designed to be a direct comparison of
the two systems, and interpreting the differences shown in Table 1 is difficult
given the many differences in cDNA/library preparation and the rapid evolution
of each system over the past year.

\subsection{Statistical analysis for differential expressions}
We used an R package, DESeq, \cite{Anders2010} to determine differentially
expressed genes based on negative binomial model originally developed by \cite{Robinson2007}.
Steps of analyzing RNA-seq data for determining differentially expressed genes
are described in \cite{Oshlack2010}. We used a short read aligner called
\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
aligner allowed a few gaps for efficient alignment of millions of sequences of
size being 100 base pairs long, short reads that consist of RNAs and sequencing
adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
remove parts of adapter sequences from short reads before mapping on the
reference genome. Mapped short read alignments were converted into readable
formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
reference genome, forming a table of read counts of the genes. Statistical
software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
was subsequently employed to infer differentially expressed in different
biological conditions.  We also generated tracks for our recently released
Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
summarize gene expression levels based the short read alignments. This allowed
us to visualize parts of the reference genome for comparing different RNA
samples. Further analyses such as functional category enrichment test
\cite{Young2010} were also employed.

\subsection{Transcript predictions}
\citet{Martin2010} developed a hidden Markov model to predict transcripts by
using site-wise expression levels. We used the following options for the
program: ``-c 10 -b 25 -force\_gp''. We binned expression levels into 25 parts.
We allowed 10 emission states. We also used the genes as the predicted parts of
transcripts. 
Because we did not have strand-specific RNA-seq of the 13 samples due to
limited resources, we could not tell in which directions transcripts based on
RNA-seq data were oriented.

\subsection{small RNAs}
RNAz is used to find small RNA candidates. Sequence alignments were a critical
step for finding small RNAs in the approach. Phylogenies of streptococcus
species could be considered to choose species related with Streptococcus mutans.
Alternatively, I find sequences similar to intergenic regions, and refine the
alignments using a multiple sequence alignment tool such as MUSCLE
\cite{Edgar2004a}. Steps for achieving this goal include: 1. Exract intergenic
regions including up- and down-stream, 2. BLAST the intergenic regions against
non-redundant DB to find similar sequences \cite{Altschul1990}, 3. Refine each
alignment using MUSCLE \cite{Edgar2004a}, 4. Apply RNAz to the alignments for
scoring regions for small RNAs \cite{Washietl2005,Gruber2010}, and 5. Find
transcripts with high score of small RNAs. The transcripts must be defined in
advance.  We used RNAplex \citep{Tafer2008} to predict targets of putative small
RNAs. The method used RNAplfold \citep{Bernhart2006}.  Please, find a review
paper about small RNA predictions and targets by \citet{Tafer2008a}.

\subsection{Gene category associations and virulence genes}
We used two approaches for finding functional categories
\citep{Young2010,Suzuki2011}. We briefly descirbe We briefly descirbe the
prodecure of \citet{Suzuki2011}.  We assigned genes to Gene Ontology (GO)
categories by comparing the {\em Streptococcus} genes to bacterial proteins from
the Uniref90 database using BLASTP, and then assigning the same GO
classification as the target gene of the uniProt GOA database if the match had
an $E$-value of $<1.0\times10^{-5}.$  A gene family was assigned a given
classification if any of its genes was assigned that classification.  To test
for associations with replacing gene transfers we used the recombination
intensity estimated for each gene, and for additive gene transfers we used the
number of additive transfers inferred for each family.  To test for
significance, we performed a Mann-Whitney $U$ test of the values (recombination
intensities or numbers of transfers) associated with a given category vs.\ the
values for the all other genes/families.  We used the \cite{Benjamini1995}
method to correct for multiple comparisons.

\section{Results}

Findings include:
\begin{enumerate}
\item New genes, open reading frames that are not annotated in NCBI's genome database.
\end{enumerate}

We sequenced a whole prokaryotic transcriptome of \textit{S\. mutans}, and
mapped short reads to the genomic sequence of the reference genome UA159
\citep{Ajdic2002}. 
\citet{Martin2010} applied a hidden Markov model to use short reads mapped to a
genomic sequence in order to parase \textit{B.\ anthracis} genome into a
sequence of transcripts. Most of annotated genes belong to the predicted
transcripts. We found discrepancies of annotated genes and their transcriptions:
1. A gene contains two transcripts, 
2. Non-coding regions contain transcripts.
\fixme{We could find more discrepancies. Try to find them}. 
Expression levels of adjacent genes within a transcript should be more
correlated to each other than those of adjacent genes belonging to separate transcripts
\citep{Martin2010}.

@@

In the data generation phase (FIG. 1a), total RNAs or mRNAs are fragmented and
converted into a library of cDNAs containing sequencing adaptors.  
In the data analysis phase (FIG. 1b), these short reads are pre-processed to
remove sequencing errors and other artefacts. The reads are subsequently
assembled to reconstruct the original RNAs and to assess their abun- dance
(‘expression counting’).

@@

\subsection{My Results}

The input RNA-seq data is a list of site-wise non-negative integers that
represents a number of reads mapped to a particular genomic position
\citet{Martin2010}; a RNA-seq ``coverage'' data. We designated a gene as an
expressed if the gene belongs to a transcript with at least C2 level.

A gene

We use the term `coverage' to designate 
We verified the genes using RNA-seq data.

>>>>

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color} 
\usepackage{setspace} 
\usepackage{rotating}
\usepackage{natbib}
\usepackage{xr}
\usepackage{../../latex/sty/genres}
\externaldocument{supplementary}

@@

\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=6,height=6}
\newcommand{\Rpackage}[1]{R package \textit{#1}}
\newcommand{\program}[1]{{\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}
\newcommand{\comment}[1]{}

@@

\author{Sang Chul Choi$^{1}$, Charles D.\ Danko$^{1}$, 
Melissa J.\ Hubisz$^{1}$, \\
Paulina Dmitrievna Pavinski Bitar$^{2}$, \\
Robert A.\ Burne$^{3}$
Michael J.\ Stanhope$^{2}$, and Adam Siepel$^{1}$}

@@

\section{Abstract}
A bacterial transcriptome from the primary etiological agent of human dental
caries, \textit{Stretptococcus mutans}, is described using deep RNA sequencing
data.  We define transcripts boundaries of open reading frames based on sequence
alignments of short reads on the reference genome of UA159 from the deep
sequencing.  Based on homologous sequences and RNA secondary structures we also
report putative small RNAs, several of which are confirmed experimentally using
both RNA sequencing data, northern blot, and qRT-PCR.  CcpA is known to play a
direct role in carbon catabolite repression (CCR).  We use the developed
transcript map, open read frame, and small RNAs to determine differentially
expressed regions between the wild-type (UA159) and CcpA-deletion (TW1) strains
in order to refine the previously established importance of CcpA regulation as
central metabolism and virulence gene expression.  We find that more genes are
differentially expressed in the comparison between wild-type and CcpA-deletion
strains that are grown in galactose than in glucose.  We also find that more
genes differentially expressed in wild-type than in CcpA-deletion strain when
comparing glucose- and galactose-grown cells.  These two findings are different
from those from microarry experiment of \cite{Abranches2008}.  We also predicted
target genes of the small RNAs that were enriched in the following functional
categories: xxx.  We confirmed that CcpA modulates the pathogenic potential of
\textit{S.\ mutans} through global control of gene expression.

\comment{
\section{Abbreviations}
\begin{enumerate}
\item CCR Carbon catabolite repression
\item CcpA ?
\item IPS intracellular polysaccharide
\item CRE catabolite responsive elements
\end{enumerate}
}

@@

Small RNAs and their targets can be predicted \citep{Tafer2008a}.
Although our RNA-seq data did not sample specifically non-coding small RNAs, we
retained small RNAs in the sample by attempting to deplete only ribosomal RNAs
from the sample. Because non-coding small RNAs and coding messenger RNAs were
mixed in the sample, we sought a way of predicting small RNAs using
\program{RNAz} developed by \citet{Gruber2010}.  \program{RNAz} used homologous
sequences and RNA secondary structures to predict putative non-coding RNAs. We
confirmed the several predicted small non-coding RNAs experimentally
\citep{Altuvia2007} including the RNA-seq data set. Because small RNAs could
be associated with CCR regulation of genes (CITE), we determined differentially expressed small
RNAs between RNA samples. We found a few small RNAs to be differentially
expressed.

Bacteria can selectively use substrates from different carbon sources. In the
presence of a preferred carbon source, catabolic system of secondary substrates
are often regulated or shutdown: called carbon catabolite repression
\citep{Gorke2008}. CCR also regulates the expression of virulence factors in many
pathogenic bacteria. A classical example is the glucose-lactose diauxie in
\textit{E.\ coli}. A significant number of genes can be subject to CCR.  CCR in
bacteria involves regulatory networks that activate or silence genes in response
to carbon source and availability.  \textit{S.\ mutans} possesses redundant
systems for CCR.  We could measure directly the expression levels of
\textit{fruA} in \textit{S.\ mutans}.  We assess the contribution of CcpA to CCR
of predicted operons.  Clearly, deletion of \textit{ccpA} resulted in
substantial relief of the CCR of \textit{fruA} expression exerted by both
glucose and fructose.

@@

\fixme{We need to quantify some uncertainty of transcript predictions.}

@@

\subsection{Bacterial strains and growth conditions}
\fixme{Burne group should be able to describe this part. The following is just
from somewhere for the purpose of writing template. Please, rewrite it.}
\textit{S.\ mutans} UA159 is a human cariogenic isolate. The whole genome is
availabe at NCBI \citep{Ajdic2002}. Strain TW1 was derived from UA159 by
replacing a part of \textit{ccpA} gene \citep{Wen2002}.  \textit{S.\ mutans}
strains were inoculated into TV medium supplemented with 0.5\% galactose or
glucose and grown to early exponential phase (OD600, 0.2).  Then fructose was
added to the medium at concentrations (0,0.05,3,5,10 mM) for 3.5h to induce
\textit{fruA} expression.

\subsection{RNA isolation}
\fixme{Burne group should be able to describe this part. The following is just
from somewhere for the purpose of writing template. Please, rewrite it.}
RNA was isolated from S. mutans (1) by using 50 ml cultures that were grown
under the desired conditions and harvested by centrifugation at $4^{\circ}$C. Pelleted
cells were resuspended in 400 ul of diethyl pyrocarbonate-treated water, 800 ul
of RNAprotect reagent (Qiagen, Inc., Chatsworth, CA) was added, and the samples
were incubated at room temperature for 5 min with vortexing for 10 s at 1-min
intervals. Cells were then pelleted, resuspended in 250 ul Tris-EDTA (50:10)
buffer, and transferred to 1.5 ml screw-cap tubes containing sterile glass beads
(average diameter, 0.1 mm; Biospec, Bartlesville, OK), 10 ul of 1\% sodium
dodecyl sulfate, and 300 ul of acid phenol-chloroform (5:1). After
centrifugation for 15 min at maximum speed at $4^{\circ}$C, RNA was further purified
using an RNeasy mini kit (Qiagen, Inc., Chatsworth, CA), including on-column
DNase digestion with RNase-free DNase (Qiagen), as recommended by the supplier.

\subsection{RNA sequencing}
\fixme{SC wrote the second paragraph based on email correspondences between PDPB and CDD.
Please rewrite it if necessary. Stanhope group should be able to describe this
part. The following first paragraph is from a paper. Do not use them or rephrase
them in any way. They are for informational purposes.}
Sequencing libraries for the Illumina GA platform were constructed by shearing
the enriched cDNA by nebulisation (35psi, 6 min) followed by end-repair with
Klenow polymerase, T4 DNA polymerase and T4 polynucleotide kinase (to blunt-end
the DNA fragments). A single 3' adenosine moiety was added to the cDNA using
Klenow exo- and dATP. The Illumina adapters (containing primer sites for
sequencing and flowcell surface annealing) were ligated onto the repaired ends
on the cDNA and gel-electrophoresis was used to separate library DNA fragments
from unligated adapters by selecting cDNA fragments between 200-250 bps in size.
Fragmentation followed by gel-electrophoresis were used to separate library DNA
fragments and size fragments were recovered following gel extraction at room
temperature to ensure representation of AT rich sequences. Ligated cDNA
fragments were recovered following gel extraction at room temperature to ensure
representation of AT rich sequences. Libraries were amplified by 18 cycles of
PCR with Phusion polymerase. Sequencing libraries were denatured with sodium
hydroxide and diluted to 3.5 pM in hybridisation buffer for loading onto a
single lane of an Illumina GA flowcell. Cluster formation, primer hybridisation
and single-end, 36 cycle sequencing were performed using proprietary reagents
according to manufacturers' recommended protocol (https://icom.illumina.com/).
The efficacy of each stage of library construction was ascertained in a quality
control step that involved measuring the adapter-cDNA on a Agilent DNA 1000
chip. A final dilution of 2 nM of the library was loaded onto the sequencing
machine.

Deep sequencing of the RNA samples were performed at the Cornell University Life
Sciences Core Laboratories Center (Ithaca, NY).  RNA extraction was performed
using a modified Qiagen RNAease protocol (Qiagen, Inc., Chatsworth, CA): cells
were pelleted, resuspended in RNA protect reagent, frozen, then resuspended in
TE, mechanically disrupted in a bead beater in the presence of glass beads and
acidic phenol. After these steps, we used the Qiagen kit to obtain total RNAs.
The RNA sample was then passed through the Ambion MicrobExpress kit twice, which
removed ribosomal RNAs only using magnetic beads.  This procedure left small
RNAs and transfer RNAs intact.  We followed the TruSeq Illumina kit to prepare
the cDNA libraries.  We did not intend to enrich short reads in either small
RNAs or transcript boundaries.

\subsection{Statistical analysis for differential expressions}
We used an R package, DESeq, \citep{Anders2010} to determine differentially
expressed genes based on negative binomial model originally developed by
\citep{Robinson2007}.  Steps of analyzing RNA-seq data for determining
differentially expressed genes are described in \cite{Oshlack2010}. We used a
short read aligner called \program{BWA} \citep{Li2009a} to map the sequenced
short reads on a reference genome, \textit{Streptococcus mutans} UA159 (NCBI
accession NC\_004350). Because the aligner allowed a few gaps for efficient
alignment of millions of sequences of size being 100 base pairs long, short
reads that consist of RNAs and sequencing adapter sequences would be not mapped.
We used \program{cutadapt} \citep{Martin2011} to remove parts of adapter
sequences from short reads before mapping on the reference genome. Mapped short
read alignments were converted into readable formats using \program{samtools}
\citep{Li2009b}. We counted short reads aligned to annotated genes in the
reference genome, forming a table of read counts of the genes. Statistical
software R \citep{R2011} with \Rpackage{DEseq} \citep{Anders2010} was
subsequently employed to infer differentially expressed in different biological
conditions.  We also generated tracks for our recently released Streptococcus
Genome Browser (http://strep-genome.bscb.cornell.edu) that summarize gene
expression levels based the short read alignments. This allowed us to visualize
parts of the reference genome for comparing different RNA samples. Further
analyses such as functional category enrichment test \cite{Young2010} were also
employed.

\subsection{Transcript predictions}
\citet{Martin2010} developed a hidden Markov model to predict transcripts by
using site-wise expression levels. We used the following options for the
program: ``-c 10 -b 25 -force\_gp'' to bin expression levels into 25 parts to
allow 10 emission states. We also used the genes as the predicted parts of
transcripts.  Because we did not have strand-specific RNA-seq of the 13 samples
due to limited resources, we could not tell in which directions transcripts
based on RNA-seq data were oriented.

\subsection{small RNAs}
We used \program{RNAz} to predict small RNA candidates. Sequence alignments were
a critical step for finding small RNAs in the approach. Phylogenies of
streptococcus species could be considered to choose species related with
Streptococcus mutans.  Alternatively, we found sequences similar to intergenic
regions, and refined the alignments using a multiple sequence alignment tool
such as MUSCLE \citep{Edgar2004a}. First, we exracted intergenic regions
including up- and down-stream. We searched non-redundant database of bacterial
genomes using BLAST \citep{Altschul1990} for similar intergenic sequences,
subsequently, refining the sequence alignments using MUSCLE \citep{Edgar2004a}.
Then, we applied \program{RNAz} to the alignments for scoring regions for small
RNAs \citep{Washietl2005,Gruber2010}.  Targets of transcripts with high scores
for small RNAs were also predicted using \program{RNAplex} \citep{Tafer2008},
and \program{RNAplfold} \citep{Bernhart2006}.  Target genes were scored for each
small RNA candidate, which was used to perform functional category enrichment
tests of targets for each small RNA candidate. We could find functional
categories that a few small RNAs might be associated with.

\subsection{Gene category associations and virulence genes}
We used two approaches for finding functional categories
\citep{Young2010,Suzuki2011}. We briefly descirbe the prodecure of
\citet{Suzuki2011}.  We assigned genes to Gene Ontology (GO) categories by
comparing the \textit{S.\ mutans} genes to bacterial proteins from the Uniref90
database using \program{BLASTP}, and then assigning the same GO classification
as the target gene of the uniProt GOA database if the match had an $E$-value of
$<1.0\times10^{-5}.$  A gene family was assigned a given classification if any
of its genes was assigned that classification.  To test for associations with
replacing gene transfers we used the recombination intensity estimated for each
gene, and for additive gene transfers we used the number of additive transfers
inferred for each family.  To test for significance, we performed a Mann-Whitney
$U$ test of the values (recombination intensities or numbers of transfers)
associated with a given category vs.\ the values for the all other
genes/families.  We used the \cite{Benjamini1995} method to correct for multiple
comparisons.
\fixme{A short description about \citep{Young2010}.}

\section{Results}
\comment{
Findings include:
\begin{enumerate}
\item New genes, open reading frames that are not annotated in NCBI's genome database.
\end{enumerate}
}
We sequenced a whole prokaryotic transcriptome of \textit{S.\ mutans}, and
mapped short reads to the genomic sequence of the reference genome UA159
\citep{Ajdic2002}. 
\citet{Martin2010} applied a hidden Markov model to use short reads mapped to a
genomic sequence in order to parase \textit{B.\ anthracis} genome into a
sequence of transcripts. Most of annotated genes belong to the predicted
transcripts. We found discrepancies of annotated genes and their transcriptions:
1. A gene contains two transcripts, 
2. Non-coding regions contain transcripts.
\fixme{We could find more discrepancies. Try to find them}. 
Expression levels of adjacent genes within a transcript should be more
correlated to each other than those of adjacent genes belonging to separate transcripts
\citep{Martin2010}.

@@

In the data generation phase (FIG. 1a), total RNAs or mRNAs are fragmented and
converted into a library of cDNAs containing sequencing adaptors.  
In the data analysis phase (FIG. 1b), these short reads are pre-processed to
remove sequencing errors and other artefacts. The reads are subsequently
assembled to reconstruct the original RNAs and to assess their abun- dance
(`expression counting').

@@

<<manuscript,echo=FALSE,results=hide>>=
# not run when smutans R package is installed
# Use this when smutans R package is being tested.
library(DESeq)
library(GenomicRanges)
library(plotrix)
source("smutans/R/class_and_slots.R")
source("smutans/R/core.R")
source("smutans/R/methods.R")
load("smutans/data/smutansData.tx.RData")
load("smutans/data/smutansData.txGenes.RData")
load("smutans/data/smutansData.txPileup.RData")
load("smutans/data/smutans.feature.genes.RData")
load("smutans/data/smutans.genes.criteria.RData")
load("smutans/data/smutans.go.genes.RData")
load("smutans/data/smutans.cat.desc.RData")
load("smutans/data/smutansGenes.RData")
@
<<manuscript,eval=FALSE,echo=FALSE,results=hide>>=
# not run without smutans R package installed
# Remove eval=FALSE when the R package is installed
library(smutans)
load(smutansData.tx)
load(smutansData.txGenes)
load(smutansData.txPileup)
load(smutans.feature.genes)
data(smutans.genes.criteria)
data(smutans.go.genes)
data(smutans.cat.desc)
data(smutansGenes)
@
<<echo=FALSE,result=hide>>=
grTx <- smutansData.tx[elementMetadata(smutansData.tx)$state > 1]
mtch2 <- findOverlaps(smutansData.txGenes,grTx)
grNoncoding <- gaps(smutansData.txGenes)
# A bug in GenomicRanges: the last segment is the while genome. I removed it
# from the noncoding ranges.
grNoncoding <- grNoncoding[-length(grNoncoding)] 
mtchNoncoding <- findOverlaps(grNoncoding,grTx)
# Expressed transcripts without annotated genes
grTxWGenes <- grTx[unique(matchMatrix(mtch2)[,2])]
grTxWOGenes <- grTx[-unique(matchMatrix(mtch2)[,2])]
# Find transcripts with genes with conflicted strands
grGenesPlus <- smutansData.txGenes[strand(smutansData.txGenes)=='+']
grGenesMnus <- smutansData.txGenes[strand(smutansData.txGenes)=='-']
mtchPlus <- findOverlaps(grGenesPlus,grTxWGenes)
mtchMnus <- findOverlaps(grGenesMnus,grTxWGenes)
txWithConflictedGenes <- intersect(matchMatrix(mtchPlus)[,2],matchMatrix(mtchMnus)[,2])
grTx2 <- grTxWGenes[-txWithConflictedGenes] 
# Lengths of 5' and 3' UTRs
mtchPlus <- findOverlaps(grGenesPlus,grTx2)
mtchMnus <- findOverlaps(grGenesMnus,grTx2)
length5UTR <- c()
length3UTR <- c()
for (i in unique(matchMatrix(mtchPlus)[,"subject"])) {
  x <- matchMatrix(mtchPlus)[matchMatrix(mtchPlus)[,"subject"]==i,"query"]
  y <- min(start(grGenesPlus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesPlus[x])) 
  length5UTR <- c(length5UTR,y)
  length3UTR <- c(length3UTR,z)
}
for (i in unique(matchMatrix(mtchMnus)[,"subject"])) {
  x <- matchMatrix(mtchMnus)[matchMatrix(mtchMnus)[,"subject"]==i,"query"]
  y <- min(start(grGenesMnus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesMnus[x])) 
  length5UTR <- c(length5UTR,z)
  length3UTR <- c(length3UTR,y)
}
# Correlation of expression levels of adjacent genes.
pileup.x <- lapply(ranges(smutansData.txGenes), function(x) seqselect(smutansData.txPileup, x))
pileup.y <- mapply(function(x,y) sum(x)/y, x=pileup.x, y=width(smutansData.txGenes))
elementMetadata(smutansData.txGenes)["score"] <- pileup.y
mtch3 <- findOverlaps(smutansData.txGenes,grTx2)
x <- rle(matchMatrix(mtch3)[,"subject"])
comp.exp.two.genes <- c()
for (i in x$values[x$lengths > 1]) {
  y <- combn(matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==i],2)
  for (j in 1:dim(y)[2]) {
    comp.exp.two.genes <- cbind(comp.exp.two.genes,as.matrix(elementMetadata(smutansData.txGenes)["score"][y[,j],]))
  }
}
# Correlation of expression levels of adjacent genes between transcripts.
comp.exp.two.genes.between.transcript <- c()
for (i in 1:length(x$values)) {
  if (i == length(x$values)) {
    break
  }
  y1 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i]]
  i2 <- i + 1
  y2 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i2]]
  l <- as.matrix(do.call(expand.grid, list(y1,y2)))
  for (j in 1:dim(l)[1]) {
    comp.exp.two.genes.between.transcript <- cbind(comp.exp.two.genes.between.transcript,as.matrix(elementMetadata(smutansData.txGenes)["score"][l[j,],]))
  }
}
# plot(log10(comp.exp.two.genes.between.transcript[1,]), log10(comp.exp.two.genes.between.transcript[2,]))
# cor(log10(comp.exp.two.genes.between.transcript[1,]), log10(comp.exp.two.genes.between.transcript[2,]))
@

\subsection{My Results}
% Statistics of transcripts, and genes.
The input RNA-seq data is a list of site-wise non-negative integers that
represents a number of short reads mapped to a particular genomic position
\citet{Martin2010}, which is called a RNA-seq ``coverage'' data. We designated a
gene as an expressed if the gene belongs to a transcript with at least C2
level.
The number of predicted expressed transcripts was
\Sexpr{sum(elementMetadata(smutansData.tx)$state > 1)}
(\Sexpr{as.integer(sum(elementMetadata(smutansData.tx)$state > 1)/length(smutansData.tx)*100)}\%)
out of 
\Sexpr{length(smutansData.tx)} 
fragments.
Of the total \Sexpr{length(smutansData.txGenes)} of genes,
\Sexpr{length(matchMatrix(mtch2)[,1])}
(\Sexpr{as.integer(length(matchMatrix(mtch2)[,1])/length(smutansData.txGenes)*100)}\%) 
were designated as expressed. 
Genes that were not designated as expressed could be due to false gene
annotations or possibility of true negative expression under the experimental
conditions. 
Figure \ref{figLength} shows the length distribution of expressed and
not-expressed genes. 
Of the \Sexpr{length(grTx)} expressed transcripts, 
\Sexpr{length(grTxWOGenes)} contained no annotated genes.
Figure \ref{figTranscriptLength} shows the length distribution of expressed
transcripts with annotated coding genes and unannotated regions.
Unannotated regions can include potential non-coding regions and new
coding genes.  Transciprts with unannotated regions appeared to be shorter than those
with coding genes. Most of the unannotated regions with length being near 360 bp
could be non-coding regions, and the rest of the unannotated regions with length
being near 1000 bp could be new coding genes. Some of the unannotated genes
could be pseudogene regions.
Because we used a RNA-seq technique without considering strand-specific, we
could not tell the strandness of the predicted transcripts. Of
the \Sexpr{length(grTx)} expressed transcripts, 
\Sexpr{length(txWithConflictedGenes)}
included annotated genes of both strands
(\Sexpr{as.integer(length(txWithConflictedGenes)/length(grTx)*100)}\%).
We filter out the transcripts with genes with conflicted both strands leaving
\Sexpr{length(grTx2)} expressed transcripts. 
Figure \ref{figUTRLength} shows length distributions of 5' and 3' UTR of the
expressed transcripts without genes with conflicted strands.
Many transcripts appeared to start and end near the boundaries of annotated genes.
Gene expression levels for adjacent genes within expressed transcripts were more
correlated than those between adjacent transcripts 
(Figure \ref{figExpressionTwoGene}).
% Question: What are the promoter scores?
